Multi-target detection with application to cryo-electron microscopy

### 2019-03-05 inverse_problems/inverse_problems.tex
sec 4.3

Under the separation condition~\eqref{eq:spacing}, the relation between autocorrelations of the observation $y$ and those of $x$ is particularly simple, as we now show. It is useful to introduce some notation: let $\vert s\vert = \sum_i s[i]$ denote the number of signal occurrences in $y$, and let
\begin{equation}
\gamma  = \frac{|s| L}{N}.
\label{eq:gamma}
\end{equation}
This $\gamma$ is the fraction of entries of $y$ occupied by signal occurrences. %Henceforth, we call it the signal density.
The separation condition  imposes $\gamma\leq\frac{L}{2L-1}\approx 1/2$.

@@

\subsection{Autocorrelations under the Poisson model}

@@

In Section~\ref{sec:proof_prop_poisson}, we prove the following expressions for the autocorrelations of $y$ under the Poisson model. As in the well-separated model, the observed autocorrelations do not depend on individual occurrences of the signal, but only on the distribution of $x$ itself, and the parameters $\gamma$ and $\sigma$: \TODO{Do these identities hold even without taking a limit? I guess so, but let's be sure.}
%
\begin{align}
	a_y^1  & =  \gamma a_{x}^1, \label{eq:mean_micrograph2} \\
	a_y^2[\ell] & = \gamma a_x^2[\ell] + \sigma^2 \delta[\ell] + (\gamma a_x^1)^2, \label{eq:ac2_micrograph2} \\
	a_y^3[\ell_1,\ell_2]  & =  \gamma a_x^3[\ell_1,\ell_2] + \sigma^2 \gamma a_x^1 \big(\delta[\ell_1]+\delta[\ell_2] +\delta[\ell_1-\ell_2] \big) \nonumber \\
						  & \qquad + (\gamma a_x^1)^3 + \gamma a_x^1 \cdot ( \gamma a_x^2[\ell_1] + \gamma a_x^2[\ell_2] + \gamma a_x^2[\ell_2-\ell_1]). \label{eq:ac3_micrograph2}
\end{align}
In the homogeneous case ($K=1$), we show in Section~\ref{sec:theory_homogeneous_poisson} that $a_y^1,a_y^2$ and $a_y^3$ identify uniquely the signal $x$, the Poisson parameter $\gamma$ and the noise level $\sigma$ for generic $x$. 
%
%From these values, one may solve for the terms $\gamma a_x^1$, $\gamma a_x^2[\ell] + \sigma^2 \ell[\ell]$, and $\gamma a_x^3[\ell_1,\ell_2]$. Indeed, recovering  $\gamma a_x^1$ and $\gamma a_x^2[\ell] + \sigma^2 \ell[\ell]$ is immediate from \eqref{eq:mean_micrograph2} and \eqref{eq:ac2_micrograph2}, and $\gamma a_x^3[\ell_1,\ell_2]$ then follows from
%%
%\begin{align}
%\gamma a_x^3[\ell_1,\ell_2]
%= a_y^3[\ell_1,\ell_2] - (\gamma a_x^1)^3
%- \gamma a_x^1 \cdot \big(a_y^2[\ell_1]  
%+ a_y^2[\ell_2] + a_y^2[\ell_1-\ell_2]- 3(\gamma a_x^1)^2 \big).
%\end{align}

@@

\TODO{Check and specify if this holds for both models}
\TODO{Did we define densities $\gamma_k$?}
\TODO{We write the section for the well-separated model, but it would be the same for the Poisson model}

In the heterogeneous model~\eqref{eq:model}, the unknowns are $K$ signals of length $L$, together with their densities $\gamma_1, \ldots, \gamma_K$ (equivalently: the distribution $\pi$ and overall density $\gamma$) and possibly the noise level $\sigma$. To estimate these parameters, we must collect at least as many independent equations. Within our framework, polynomial equations are provided by the observable autocorrelations, which correspond to mixed autocorrelations of the unknowns as per~\eqref{eq:mixedautocorr}. In this section, following~\cite{boumal2017heterogeneous}, we count how many equations the first three autocorrelations may provide in the best case (discounting symmetries). This leads to a straightforward information-theoretic upper bound on the number $K$ of signals which can be estimated, as a function of $L$. This is only an upper bound, though a bound of the same type was shown to be tight in a similar setting~\cite{bandeira2017estimation}.

@@

Hence, if $\sigma$ is known we collect
\begin{align*}
	1 + L + \frac{(L+1)(L+2)}{2}-2 = \frac{1}{2} L (L+5)
\end{align*}
equations, while if it unknown and we choose not to estimate it, then we collect
\begin{align*}
	1 + (L-1) + \frac{(L-1)(L-2)}{2} = \frac{1}{2} L (L-1) + 1
\end{align*}
equations in total. Of course, there may be redundancy in these equations: we aim only to provide a bound, under the assumption that these equations are independent.

Since we aim to estimate $KL$ parameters for the $K$ signals of length $L$ plus $K$ parameters for the densities $\gamma_k$ (or the distribution $\pi$ and overall density $\gamma$), there are $K(L+1)$ unknowns. As a result, an absolute upper bound on $K$ such that the estimation problem may be solvable is
\begin{align*}
	K \leq \frac{L(L+5)}{2(L+1)},
\end{align*}
for the case of $\sigma$ known, and
\begin{align*}
	K \leq \frac{ L (L-1) + 1}{2(L+1)}
\end{align*}
for the case of $\sigma$ unknown and not estimated. Overall, this indicates that, at best, approximately $L/2$ signals and their densities can be recovered from the first three mixed autocorrelations. Based on related results in~\cite{bandeira2017estimation}, we expect that as many as $L/2$ signals can indeed be estimated, though possibly not with computationally tractable estimators.

>>>>

Under the separation condition~\eqref{eq:spacing}, the relation between autocorrelations of the observation $y$ and those of $x$ is particularly simple, as we now show. It is useful to introduce some notation: let $\vert s\vert = \sum_i s[i]$ denote the number of signal occurrences in $y$, and let
\begin{equation}
\gamma  = \frac{|s| L}{N}.
\label{eq:gamma}
\end{equation}
This $\gamma$ is the fraction of entries of $y$ occupied by signal occurrences. %Henceforth, we call it the signal density.
The separation condition  imposes $\gamma\leq\frac{L}{2L-1}\approx 1/2$. 
For the heterogeneous model~\eqref{eq:model},  one can define
\begin{equation}
\gamma_k  = \frac{|s_k| L}{N},
\label{eq:gamma_k}
\end{equation}
where $\vert s_k\vert = \sum_i s_k[i]$ so that $\gamma=\sum_k \gamma_k$.

@@

\subsection{Autocorrelations under the Poisson model} \label{sec:ac_poisson}

@@

In Section~\ref{sec:proof_prop_poisson}, we prove the following expressions for the autocorrelations of $y$ under the Poisson model. As in the well-separated model, the observed autocorrelations do not depend on individual occurrences of the signal, but only on the distribution of $x$ itself, and the parameters $\gamma$ and $\sigma$: \TODO{Do these identities hold even without taking a limit? I guess so, but let's be sure.}
%
\begin{align}
	a_y^1  & =  \gamma a_{x}^1, \label{eq:mean_micrograph2} \\
	a_y^2[\ell] & = \gamma a_x^2[\ell] + \sigma^2 \delta[\ell] + (\gamma a_x^1)^2, \label{eq:ac2_micrograph2} \\
	a_y^3[\ell_1,\ell_2]  & =  \gamma a_x^3[\ell_1,\ell_2] + \sigma^2 \gamma a_x^1 \big(\delta[\ell_1]+\delta[\ell_2] +\delta[\ell_1-\ell_2] \big) \nonumber \\
						  & \qquad + (\gamma a_x^1)^3 + \gamma a_x^1 \cdot ( \gamma a_x^2[\ell_1] + \gamma a_x^2[\ell_2] + \gamma a_x^2[\ell_2-\ell_1]). \label{eq:ac3_micrograph2}
\end{align}
Note that from those autocorrelations, it is easy to retrieve the autocorrelations of the well-separated model for all entries unaffected by $\sigma$. If $\sigma$ is known, then it is true for all entries. \TODO{addition}

In the homogeneous case ($K=1$), we show in Section~\ref{sec:theory_homogeneous_poisson} that $a_y^1,a_y^2$ and $a_y^3$ identify uniquely the signal $x$, the Poisson parameter $\gamma$ and the noise level $\sigma$ for generic $x$.

%
%From these values, one may solve for the terms $\gamma a_x^1$, $\gamma a_x^2[\ell] + \sigma^2 \ell[\ell]$, and $\gamma a_x^3[\ell_1,\ell_2]$. Indeed, recovering  $\gamma a_x^1$ and $\gamma a_x^2[\ell] + \sigma^2 \ell[\ell]$ is immediate from \eqref{eq:mean_micrograph2} and \eqref{eq:ac2_micrograph2}, and $\gamma a_x^3[\ell_1,\ell_2]$ then follows from
%%
%\begin{align}
%\gamma a_x^3[\ell_1,\ell_2]
%= a_y^3[\ell_1,\ell_2] - (\gamma a_x^1)^3
%- \gamma a_x^1 \cdot \big(a_y^2[\ell_1]  
%+ a_y^2[\ell_2] + a_y^2[\ell_1-\ell_2]- 3(\gamma a_x^1)^2 \big).
%\end{align}

@@

\TODO{Check and specify if this holds for both models: see below}

\TODO{Did we define densities $\gamma_k$? TB: yes, in section 3}

In the heterogeneous model~\eqref{eq:model}, the unknowns are $K$ signals of length $L$, together with their densities $\gamma_1, \ldots, \gamma_K$~\eqref{eq:gamma_k} (equivalently: the distribution $\pi$ and overall density $\gamma$) and possibly the noise level $\sigma$. To estimate these parameters, we must collect at least as many independent equations. Within our framework, polynomial equations are provided by the observable autocorrelations, which correspond to mixed autocorrelations of the unknowns as per~\eqref{eq:mixedautocorr}. In this section, following~\cite{boumal2017heterogeneous}, we count how many equations the first three autocorrelations may provide in the best case (discounting symmetries). This leads to a straightforward information-theoretic upper bound on the number $K$ of signals which can be estimated, as a function of $L$. This is only an upper bound, though a bound of the same type was shown to be tight in a similar setting~\cite{bandeira2017estimation}.
The counting is based on the autocorrelations of the well-separated model. For entries independent of $\sigma$, the autocorrelations of the Poisson process are equivalent to those of the well-separated model. If $\sigma$ is known, it holds true for all entries; see Section~\ref{sec:ac_poisson}.

@@

Hence, if $\sigma$ is known we collect
\begin{align*}
	1 + L + \frac{(L+1)(L+2)}{2}-2 = \frac{1}{2} L (L+5)
\end{align*}
equations, while if it unknown and we choose not to estimate it, then we collect
\begin{align*}
	1 + (L-1) + \frac{(L-1)(L-2)}{2} = \frac{1}{2} L (L-1) + 1
\end{align*}
equations in total. Of course, there may be redundancy in these equations: we aim only to provide an upper bound. %under the assumption that these equations are independent.

Since we aim to estimate $KL$ parameters for the $K$ signals of length $L$ plus $K$ parameters for the densities $\gamma_k$, %(or the distribution $\pi$ and overall density $\gamma$)
 there are $K(L+1)$ unknowns. As a result, an absolute upper bound on $K$ such that the estimation problem may be solvable is
\begin{align*}
	K \leq \frac{L(L+5)}{2(L+1)},
\end{align*}
for the case of $\sigma$ known, and
\begin{align*}
	K \leq \frac{ L (L-1) + 1}{2(L+1)}
\end{align*}
for the case of $\sigma$ unknown and not estimated. Overall, this indicates that, at best, approximately $L/2$ signals and their densities can be recovered from the first three mixed autocorrelations. Based on related results in~\cite{bandeira2017estimation}, we expect that as many as $L/2$ signals can indeed be estimated, though possibly not with computationally tractable estimators.
